---
title: "Census Project"
output: pdf_document

---
1. Load requied libraries.
Install package install.packages("caret")
Install package install.packages("corrplot")
Install package install.packages('Boruta')


```{r library}
library(ggplot2)
library(corrplot)
library(Boruta)
library(caret)
```


2. Load census data.
```{r census}

setwd("c:/Ryerson University/Semester 4/ProjectCode")
loc<-getwd()
censusdata <- read.csv(file="census.csv",header=TRUE,sep=",", na.string = "?")

```



2.1. Divide the data into train and test data. 
```{r}
inTrain <- createDataPartition(y=censusdata$income, p= 0.75, list=FALSE)
training <- censusdata[inTrain,]
testing <- censusdata[-inTrain,]
```


3. Display dimensions, summary of data, names and structure of data.

```{r }
data <- training
dim(data)
nrow(data)
ncol(data)
dim(testing)
summary(data)
names(data)
str(data)
```


4. Display Class Distributions.
```{r}
# Imbalance data 
result = summary(data$income)/nrow(data) * 100
ggplot(data=data,aes(income)) + geom_bar(aes(fill = income), color = "black")
result 
```


5. Check and Cleaning missing values.

```{r}
cat("Number of missing values in training set is:", sum(is.na(data)), "\n")
na_count <-sapply(data, function(y) sum(length(which(is.na(y)))))
na_count <- data.frame(na_count)
na_count
nrow(data)
data <- na.omit(data)
nrow(data)
nrow(testing)
cat("Number of missing values in test set is:", sum(is.na(testing)), "\n")
na_count1 <-sapply(testing, function(y) sum(length(which(is.na(y)))))
na_count1
testingdata <- na.omit(testing)
nrow(testingdata)

```
5.1 Re-factoring the work class, occupation and native country after removing the NA values (exclude levels not required).

```{r}

data$workclass <- factor(data$workclass)
data$occupation <- factor(data$occupation)
data$native.country <- factor(data$native.country)
```

5.1 Re-factoring the work class, occupation and native country after removing the NA values (exclude levels not required) for testing data also.
```{r}

testingdata$workclass <- factor(testingdata$workclass)
testingdata$occupation <- factor(testingdata$occupation)
testingdata$native.country <- factor(testingdata$native.country)
```



6. Statistics of Numerical attributes

```{r}
# statistics of numerical attributes
summary(data$age)
sd(data$age)
hist(data$age, main = "Distribution of Age",xlab = "Age of an Individual" ,col ="blue")
boxplot(data$age,main="Age ")
summary(data$education.num)
sd(data$education.num)
hist(data$education.num,main = "Distribution of Education in years",xlab="Education in Years (yrs)",col = "blue")
boxplot(data$education.num,main="Distribution of Education")
summary(data$capital.gain)
sd(data$capital.gain)
hist(data$capital.gain,main = "Distribution of Capital Gain",xlab="Capital Gain",col = "blue")
boxplot(data$capital.gain,main="Capital Gain")
summary(data$capital.loss)
sd(data$capital.loss)
hist(log10(data$capital.loss),main = "Distribution of Capital Loss",xlab="Capital Loss",col = "blue")
boxplot(data$capital.loss,main="Capital Loss")
summary(data$hours.per.week)
sd(data$hours.per.week)
hist(data$hours.per.week,main = "Distribution of Hours Worked per Week",xlab="Hours worked per week",col = "blue")
boxplot(data$hours.per.week,main="Hours Worked per Week")

```

7. Correlation between numerical attributes.

```{r}
#Changing income to 0 <= 50k, 1 > 50k

data1 <- data
data1$income <- as.numeric(data1$income)-1
#Correlation plot
M <- c(1, 3, 5, 11:13, 15)
corrplot(cor(data1[,M]),method = "number")

##########################################################
# Correlations shows that numeric attributes are related #
# but are not strongly correlated. The variables are positively
# correlated. Education has the highest correlation 0.33 with income.
# Capital gain 0.22, age 0.24 and hours worked 0.23. The variables are positively correlted 
# with each other.
##########################################################
```

8. Exploratory analysis of the attribute native country.

```{r}
summary(data$native.country)

```

9.1 Reducing/Combining  levels of native country in training data.

```{r}
data$native.country <- as.character(data$native.country)
asia <- c("Cambodia", "China", "Hong", "India", "Iran", "Japan", "Laos","Philippines", "Taiwan", "Thailand", "Vietnam")
northAmerica <- c("Canada", "Cuba", "Dominican-Republic", "El-Salvador", "Guatemala","Haiti", "Honduras", "Jamaica", "Mexico", "Nicaragua","Outlying-US(Guam-USVI-etc)", "Puerto-Rico", "Trinadad&Tobago","United-States")
southAmerica <- c("Columbia", "Ecuador", "Peru")
europe <- c("England", "France", "Germany", "Greece", "Holand-Netherlands","Hungary", "Ireland", "Italy", "Poland", "Portugal", "Scotland",
            "Yugoslavia")
other <- c("South")
data$native.country[data$native.country %in% northAmerica] <- "North America"
data$native.country[data$native.country %in% asia] <- "Asia"
data$native.country[data$native.country %in% southAmerica] <- "South America"
data$native.country[data$native.country %in% europe] <- "Europe"
data$native.country[data$native.country %in% other] <- "Other"
table(data$native.country)
data$native.country <- as.factor(data$native.country)
levels(data$native.country)
## ----------------------------------------------------------------------------------
## Reduce the level of native country into 5 levels "Asia", "Europe", "North America"
## "Other", "South America"
## ----------------------------------------------------------------------------------

ggplot(data, aes(x=data$native.country,fill=data$income)) + geom_bar(position = "stack", color = "black") + theme(axis.text.x=element_text(angle = 70 , hjust= 1, size=7)) + scale_fill_brewer(palette="Paired")
## ----------------------------------------------------------------------------------
## Native conutry of the Majority of the population is North America. 
## ----------------------------------------------------------------------------------
```

9.2. Reducing/Combining  levels of native country in testing data.

```{r}
testingdata$native.country <- as.character(testingdata$native.country)
testingdata$native.country[testingdata$native.country %in% northAmerica] <- "North America"
testingdata$native.country[testingdata$native.country %in% asia] <- "Asia"
testingdata$native.country[testingdata$native.country %in% southAmerica] <- "South America"
testingdata$native.country[testingdata$native.country %in% europe] <- "Europe"
testingdata$native.country[testingdata$native.country %in% other] <- "Other"
table(testingdata$native.country)
testingdata$native.country <- as.factor(testingdata$native.country)
levels(testingdata$native.country)

```


10.1. Combining categories of work class in training data.

```{r}
ggplot(data, aes(x=data$workclass,fill=data$income)) + geom_bar(position = "stack", color = "black") + ggtitle('     Income levels in different Work Class')+  theme(axis.text.x=element_text(angle = 70 , hjust= 1, size=7))  + scale_fill_brewer(palette="Paired")

data$workclass <- gsub('^Federal-gov', 'Government', data$workclass)
data$workclass <- gsub('^Local-gov', 'Government', data$workclass)
data$workclass <- gsub('^State-gov', 'Government', data$workclass) 


data$workclass <- gsub('^Self-emp-inc', 'Self-Employed', data$workclass)
data$workclass <- gsub('^Self-emp-not-inc', 'Self-Employed', data$workclass)

data$workclass <- gsub('^Other', 'Other', data$workclass)
data$workclass <- gsub('^Unknown', 'Other', data$workclass)

data$workclass <- as.factor(data$workclass)
ggplot(data, aes(x=data$workclass,fill=data$income)) + geom_bar(position = "stack", color = "black") + ggtitle('    Income Levels in different Work Class')+ theme(axis.text.x=element_text(angle = 70 , hjust= 1, size=7))  + scale_fill_brewer(palette="Paired")
## ----------------------------------------------------------------
## Replace Federal-gov, Local_gov and State_gov into government.
## Self-emp-inc and self-emp-not-inc into Self-Employed.
## other and unknown into other.
##  --------------------------------------------------------------

## Observations: Most of the people earning more than 50K are in private sector
## after that self employment and then Government. 

```

10.2. Combining categories of work class in testing data.

```{r}
testingdata$workclass <- gsub('^Federal-gov', 'Government', testingdata$workclass)
testingdata$workclass <- gsub('^Local-gov', 'Government', testingdata$workclass)
testingdata$workclass <- gsub('^State-gov', 'Government', testingdata$workclass) 


testingdata$workclass <- gsub('^Self-emp-inc', 'Self-Employed', testingdata$workclass)
testingdata$workclass <- gsub('^Self-emp-not-inc', 'Self-Employed', testingdata$workclass)

testingdata$workclass <- gsub('^Other', 'Other', testingdata$workclass)
testingdata$workclass <- gsub('^Unknown', 'Other', testingdata$workclass)
testingdata$workclass <- as.factor(testingdata$workclass)
```


11. Relationship between categorical variables and income.
```{r}

ggplot(data, aes(x=data$education,fill=data$income)) + geom_bar(position = "stack", color = "black") + theme(axis.text.x=element_text(angle = 70 , hjust= 1, size=7)) + scale_fill_brewer(palette="Paired")

##-----------------------------------------------------------------------------
## The plot shows that the maximum number of adults earning income greater than 50K 
## have bachelor's degree. 
## In doctorate and masters also, the largest proportion is earning greater than 50 K. 
## In lower  education levels the largest proportion have income less than 50K. 
## Higher education results in higher income. 
##-----------------------------------------------------------------------------


ggplot(data, aes(x=data$sex,fill=data$income)) + geom_bar(position = "stack", color = "black") + theme(axis.text.x=element_text(angle = 70 , hjust= 1, size=7)) + scale_fill_brewer(palette="Paired")
##---------------------------------------------------------------------------
## Ratio of Males earning income greater than 50K are more as compare to female.
## ---------------------------------------------------------------------------


ggplot(data, aes(x=data$race,fill=data$income)) + geom_bar(position = "stack", color = "black") + theme(axis.text.x=element_text(angle = 70 , hjust= 1, size=7)) + scale_fill_brewer(palette="Paired")
  

##-----------------------------------------------------------------------------
## Observations: Plot shows that in terms of race the highest earning people are  
## from race White,then Asian-pacific and black.
##-----------------------------------------------------------------------------

ggplot(data, aes(x=data$marital.status,fill=data$income)) + geom_bar(position = "stack", color = "black") + theme(axis.text.x=element_text(angle = 70 , hjust= 1, size=7)) + scale_fill_brewer(palette="Paired")

ggplot(data, aes(x=data$relationship,fill=data$income)) + geom_bar(position = "stack", color = "black") + theme(axis.text.x=element_text(angle = 70 , hjust= 1, size=7)) + scale_fill_brewer(palette="Paired")

##-----------------------------------------------------------------------------
## Observations: Plots shows that married people are earning more than 50k.
##-----------------------------------------------------------------------------

ggplot(data, aes(x=data$occupation,fill=data$income)) + geom_bar(position = "stack", color = "black") + theme(axis.text.x=element_text(angle = 70 , hjust= 1, size=7)) + scale_fill_brewer(palette="Paired")

##-----------------------------------------------------------------------------
## Observations: Plots shows that in terms of occuupation people with managerical job and 
## professors are earning more than 50 K in the highest ratio. 
## Showing that people at highest post are earning more.  
##-----------------------------------------------------------------------------

```

Saving the clean test and train data  in testdata.csv and traindata.csv respectively.
```{r}
write.csv(data, "traindata.csv", row.names = FALSE)
write.csv(testingdata, "testdata.csv", row.names = FALSE)

```





